Enzo+Moray: Radiation Hydrodynamics Adaptive Mesh Refinement Simulations with Adaptive Ray Tracing

### 2011-02-02 mnras/mnras.tex
Started to respond to the referee report.

In this paper, we present our implementation, \moray, of adaptive ray
tracing \citep{Abel02_RT} in the cosmological hydrodynamics adaptive
mesh refinement (AMR) code, \enzo~\citep{BryanNorman1997, OShea2004}.
The radiation field is coupled to the hydrodynamics solver at small
time-scales, enabling it to study radiation hydrodynamical problems.
We have used this code to investigate the growth of an \hii region
from a 100\Ms~Population III (Pop III) star \citep{Abel07}, the early
stages of reionisation from Pop III stars \citep{Wise08_Reion},
radiative feedback on the formation of high redshift dwarf galaxies
\citep{Wise08_Gal, Wise10_Gal}, ultraviolet radiation escape fractions
from dwarf galaxies before reionisation \citep{Wise09}, negative
radiative feedback from accreting Pop III seed black holes
\citep{Alvarez09}, and radiative feedback in accreting supermassive
black holes \citep[][in prep.]{Kim11}.  We have included \moray in the
latest public release of \enzo\footnote{http://enzo.googlecode.com},
and it is also coupled with the newly added MHD solver in
\enzo~\citep{Wang09}.

We have structured this paper as follows.  In Section 2, we describe
the mathematical connections between adaptive ray tracing and the
radiative transfer equation.  Furthermore, we detail how physics other
than photo-ionisation and photo-heated are included.  We then derive a
geometric correction factor to any ray tracing method to improve
accuracy.  We end the section by describing a new computational
technique to approximate an optically-thin radiation field with ray
tracing and multiple sources.  In Section 3, we cover the details of
our radiation hydrodynamics implementation in \enzo, specifically (1)
the ray tracing algorithms, (2) coupling with the hydrodynamics
solver, (3) several methods to calculate the radiative transfer
timestep, and (4) our parallelisation strategy.  We present our
results from the RT06 radiative transfer tests in Section 4.
Afterwards in Section 5, we show the results from the RT09 radiation
hydrodynamics tests.  We expand on these tests to include more
dynamical and complex setups to demonstrate the flexibility and high
fidelity of \moray.  Section 6 gives the results from spatial,
angular, frequency, and temporal resolution tests.  In Section 7, we
illustrate the improvements from the geometric correction factor and
our optically-thin approximation.  We also show the effects of X-ray
radiation and radiation pressure in this section.  Finally in Section
8, we demonstrate the parallel scalability of \moray.  Last Section 9
summarises our method and results.

@@

A ray propagates and splits until 
%
\begin{enumerate}
\item the photon has travelled $c \times dt_P$, where $dt_P$ is the
  radiative transfer timestep,
\item its photon flux is almost fully absorbed ($>99.9\%$) in a single
  cell, which significantly reduces the computational time if the
  radiation volume filling fraction is small,
\item the photon leaves the computational domain with isolated
  boundary conditions, or
\item the photon travels $\sqrt{3}$ of the simulation box length with
  periodic boundary conditions.
\end{enumerate}
%
In the first case, the photon is halted at that position and saved,
where it will be considered in the solution of $I_\nu$ at the next
timestep.  In the next timestep, the photon will encounter a different
hydrodynamical and ionisation state, hence $\kappa$, in its path.
Furthermore any time variations of the luminosities will be retained
in the radiation field.  This is how this method retains the time
derivative of the radiative transfer equation.

@@

Ionisation front velocities can approach significant fractions of the
speed of light in steep density gradients and in the early expansion
of the \hii region.  If the ionisation front position is
critical to the calculation, the radiation transport timestep can be
derived from an estimate of the ionisation front velocity
%
\begin{equation}
  \label{eqn:dt_incident}
  v_{\rm IF}(\mathbf{r}) \approx \frac{F(\mathbf{r})}
  {n_{\rm abs}(\mathbf{r})},
\end{equation}
%
based on the incident radiation field at a particular position.
Alternatively, the propagation of the ionisation front can be
restricted by limiting the change in specific intensity $I$ to a
safety factor $C_{\rm RT,cfl}$, resulting in a timestep of
%
\begin{equation}
  \label{eqn:vifront1}
  dt_P = C_{\rm RT,cfl} \frac{I}{|dI/dt|}.
\end{equation}
%
We consider the specific intensity after the ray travels through the
cell, so $I = I_0 \exp(-\tau)$, where $\tau = n_{\rm H} \sigma dl$ is
the optical depth through the cell.  The time derivative of $I$ is
%
\begin{equation}
  \label{eqn:vifront2}
  \frac{dI}{dt} = I_0 \exp(-\tau) (-\dot{n}_{\rm H} \sigma dl),
\end{equation}
%
which can be expressed in terms of local optical depth and neutral
fraction,
\begin{equation}
  \label{eqn:vifront3}
    \frac{dI}{dt} = -I \frac{\tau \dot{n}_{\rm H}}{n_{\rm H}}.
\end{equation}
%
$\dot{n}_{\rm H}$ is computed with the same formula as equation
(\ref{eqn:dtelec}). Substituting in equation (\ref{eqn:vifront1})
gives
%
\begin{equation}
  \label{eqn:vifront4}
  dt_P = C_{\rm RT,cfl} \frac{n_{\rm H}}{\tau \dot{n}_{\rm H}}.
\end{equation}
%
In practice, we have found that a ceiling of 3 can be placed on the
optical depth, so optically thick cells do not create a very small
timestep.  We still find excellent agreement with analytical solutions
with this approximation.  We show the accuracy using this timestep
method in \S\ref{sec:dt_dep}.

@@

In Fig. \ref{fig:test1_ifront}, we show (a) the evolution of the
neutral and ionisation fraction as a function of radius at $t = $ 10,
30, 100, and 500 Myr, and (b) the growth of the ionisation front
radius as a function of time and its ratio with the analytical
Str\"{o}mgren radius [equation (\ref{eqn:rstr})].  The ionisation
front has a width of $\sim0.7$ kpc, which is in agreement with the
inherent thickness of $\sim18 \lambda_{\rm mfp} = 0.74$ kpc, given a
13.6 eV mono-chromatic spectrum.  There are small kinks in the neutral
fraction at 1.5 and 3 kpc that corresponds to artifacts created by the
photon package splitting at these radii.  However these do not affect
the overall solution.  One difference between our results and the
codes presented in RT06 is the increasing neutral fraction outside of
the \hii.  This occurs because the initial ionised fraction and
temperature is set to $1.3 \times 10^{-3}$ and 8000 K, which are not
the equilibrium values.  Over the 500 Myr in the calculation, the
neutral fraction increases to 0.2, which is close to its equilibrium
value.  In the right panel of Fig. \ref{fig:test1_ifront}, the
ionisation front radius exceeds $R_s$ by a few percent for most of the
calculation.  This difference happens because the analytical solution
[equation (\ref{eqn:rstr})] assumes the \hii region has a constant
ionised fraction.  The evolution of the ionised fraction as a function
of radius can be analytically calculated \citep[e.g.][]{Osterbrock89,
  Petkova09}, causing the ionisation front radius to be slightly
larger, increasing from 0 to 3\% in the interval 80--350 Myr.  Our
results are in excellent agreement with this more accurate analytical
solution.  In Fig. \ref{fig:test1_HI}, we show a slice of the neutral
fraction through the origin.  Other than the ray splitting artifacts
that generate the plateaus at 1.5 and 3 kpc, one sees spherical
symmetry without any noise in our solution.

@@

We have completed presenting results from the RT06 and RT09 test
suites.  Next we test \moray with five additional tests that further
demonstrate its capabilities and accuracy.

\subsection{Test 8. Champagne flow from a dense clump}

\begin{figure*}
  \epsscale{2}
  \plotone{f25}
  \epsscale{1}
  \caption{\label{fig:test8_1} Test 8. (Champagne flow from a dense
    clump).  Slices of density through the initial clump centre in the
    x-y plane (top) and x-z plane (bottom) at $t = 10, 40, 100, 150$
    kyr.  Notice the instabilities that grow from perturbations
    created while the \hii region is contained in the dense
    clump.}
\end{figure*}

@@

\subsection{Test 9. Irradiated Rayleigh-Taylor instability}

\begin{figure}
  %\epsscale{0.75}
  \plotone{f26a}
  \plotone{f26b}
  \plotone{f26c}
  \caption{\label{fig:test9_thick} Test 9. (Irradiated Rayleigh-Taylor
    instability; optically thick case).  Slices at $y=0$ of density
    (top), temperature (middle), and electron fraction (bottom).  The
    source turns on at $t=0$.}
  %\epsscale{1}
\end{figure}

Here we combine the classic case of a Rayleigh-Taylor instability and
an expanding \hii region.  The Rayleigh-Taylor instability
occurs when a dense fluid is being supported by a lighter fluid,
initially in hydrostatic equilibrium, in the presence of a constant
acceleration field.  This classic test alone evaluates how subsonic
perturbations evolve.  We consider the case of a single-mode
perturbation.  The system evolves without any radiation until the
perturbation grows considerably and then turn on the radiation source.
These tests demonstrate that \moray can follow a highly dynamic
system and resolve fine density structures.

@@

\subsection{Test 10. Photo-evaporation of a blastwave}

\begin{figure*}
  \epsscale{2}
  \plotone{f28}
  \epsscale{1}
  \caption{\label{fig:test10} Test 10 (Photo-evaporation of a
    blastwave).  Slices of density (top) and temperature (bottom) at
    $t = 2.5, 3, 5, 7.5$ Myr in the $x-z$ plane.  As the R-type
    ionisation front propagates through the blastwave centre,
    instabilities grow from the slightly inhomogeneous hot and rarefied
    medium.  Note that the dense shell of the blastwave also creates
    dense inward fingers in the ionisation front shock.}
  %\epsscale{1}
\end{figure*}

@@

\subsection{Test 11. Collimated radiation from a dense clump}

\begin{figure*}
  %\epsscale{1}
  \epsscale{2}
  \plotone{f29}
  \epsscale{1}
  \caption{\label{fig:test11} Test 11 (Collimated radiation from a
    dense clump).  Slices of density (top) and temperature (bottom) at
    $t = 0.1, 3.25, 10.75, 23.25$ Myr.  The conical \hii region
    drives shocks transversely into the overdense sphere and creates
    polar champagne flows.  The ambient medium is heated to $T \sim 3
    \times 10^4$ K as the ionisation front passes the
    constant-pressure cloud surface.  The ionisation front changes
    from D-type to R-type after it enters the ambient medium.}
  %\epsscale{1}
\end{figure*}

@@

We use a domain that is 2 kpc wide and has an ambient medium with
$\rho_0 = 10^{-3}$ \cubecm, $T = 10^4$ K, $x_e = 0.99$.  We place a
dense clump with $\rho/\rho_0 = 100$, $T = 100$ K, $x_e = 10^{-3}$,
and $r = 250$ pc, at the centre of the box.  Radiation is emitted in
two polar cones with $\theta_c = \pi/6$ with 768 (HEALPix level 3)
initial rays, a total luminosity of $10^{49}$ erg s$^{-1}$, and a 17.6
eV mono-chromatic spectrum.  This results in $t_{\rm rec} = 1.22$ Myr
and $R_s = 315$ pc, just outside of the sphere.  The base grid has a
resolution of 64$^3$, and it is refined with the same overdensity
criterion as Test 8.  We run this test for 25 Myr.

@@

\subsection{Test 12. Time variations of the source luminosity}

\begin{figure}
  %\epsscale{1.15}
  \plottwo{f30a}{f30b}
  \caption{\label{fig:test12_kph} Test 12 (Time variations of the
    source luminosity).  Top: Slice of the photo-ionisation rate
    \kph~through the origin.  The source has a duty cycle of 0.5 Myr,
    and the box has a light crossing time of 3.3 Myr.  The shells of
    high \kph~originate from radiation that was emitted when the
    source was at its peak luminosity, illustrating the
    time-dependence of the radiative transfer equation.  Bottom:
    Radial profile \kph~with the inverse square law overplotted.}
  %\epsscale{1}
\end{figure}

\begin{figure}
  %\epsscale{0.8}
  \plotone{f31}
  \caption{\label{fig:test12_profiles} Test 12 (Time variations of the
    source luminosity).  Radial profiles of (clockwise from upper
    left) density, temperature, pressure, and ionised fraction at $t =
    50, 100, 150, 200, 250$ Myr.  In this problem, the time variations
    in the source have little effect on the overall \hii region
    expansion.  Inside the ionisation front at $t = 50$ Myr, there are
    small density perturbations that are created by the variable
    source that are later smoothed out over a sound crossing time.}
  %\epsscale{1}
\end{figure}

\begin{figure*}
  \epsscale{2}
  \plotone{f32}
  \epsscale{1}
  \caption{\label{fig:test14_1} Test 14. (\hii region with MHD).
    Left to right: slices of density at $t = 0.18, 0.53, 1.58$ Myr in
    the x-y plane.  The streamlines show the magnetic field.}
\end{figure*}

\begin{figure*}
  \epsscale{2}
  \plotone{f33}
  \epsscale{1}
  \caption{\label{fig:test14_2} Test 14. (\hii region with
    MHD). Slices of density (top) and the x-component of the magnetic
    field (bottom) in the y-z plane at $t = 0.18, 0.53, 1.58$ Myr
    (left to right).}
\end{figure*}

@@

\subsection{Test 14. \hii region with MHD}

@@

Last we demonstrate the parallel performance of \moray in weak and
strong scaling tests.  For large simulations to consider radiative
transfer, it is imperative that the code scales to large number of
processors.

@@

Weak scaling tests demonstrates how the code scales with the number of
processors with a constant amount of work per processor.  Here we
construct a test problem with a $64^3$ block per core.  The grid is
not adaptively refined.  The physical setup of the problem is nearly
the same as Test 5 with a uniform density $\rho = 10^{-3} \;\cubecm$
and initial temperature $T = 100$ K.  Each block has the same size of
15 kpc as Test 5.  At the centre of each grid, there exists a
radiation source with a luminosity $L = 5 \times 10^{48}$ ph s$^{-1}$
and a 17 eV mono-chromatic spectrum.  The problem is run for 250 Myr.
We run this test with $N_p = 2^n$ cores with $n = [0,1 \dots 10,11]$.
The domain has $(N_x, N_y, N_z)$ blocks that is determined with the
MPI routine \texttt{MPI\_Dims\_create}.  For example with $n = 7$, the
problem is decomposed into $(N_x, N_y, N_z) = (4,4,8)$ blocks,
producing a $256 \times 256 \times 512$ grid.  We have run these on
the original (Harpertown CPUs) nodes of the NASA NAS machine,
Pleiades, with 8 cores per node.  Fig. \ref{fig:weak} shows the
performance timings of various parts of the code.  From one to two
cores, the total time only increases by $\sim$1\% due to the overhead
associated with the inter-processor message passing.  From two to
eight cores, the performance decreases in the hydrodynamics solver,
chemistry and energy solver, and obtaining the hydrodynamic boundary
conditions because of cache locality problems of the data being passed
to the CPU.  This occurs because of the CPU architecture, specifically
the L1 cache and core connectivity.  We see less of a penalty in newer
processors, e.g. Intel Nehalam and Westmere CPUs.  Above eight
processes, these routines exhibit near perfect weak scaling to 2048
cores.  Unfortunately there exists a $N_p^{1.5}$ dependence in the ray
communication routines.  It becomes the dominant process above 512
processes.  We are actively pursing a solution to this scaling
problem.  The other parts of the code exhibit excellent weak scaling.
Overall it scales already well to 512 processes, and there is room to
enhance the weak scalability of \moray in the near future.

@@

Strong scaling tests shows how the problem scales with the number of
processors for the same problem.  The overhead associated with the
structured AMR framework in \enzo~can limit the strong scalability.
One key property of strong scaling is that each processor must have
sufficient work to compute, compared to the communication involved.
In our experience, non-AMR calculations exhibit much better strong
scaling than AMR ones because of reduced inter-processor
communication.  We use an AMR simulation to demonstrate the scalability
of \moray in a demanding, research application.  Here we use a
small-box reionisation calculation with $L_{\rm box} = 3$ Mpc/h, a
resolution of $256^3$, and six levels of refinement.  We measure the
time spent on the hydrodynamics, non-equilibrium chemistry, ray
tracing, and radiation transport communication in a timestep lasting 1
Myr, at $z=10$.  There are nine radiative transfer timesteps in this
period.  The box has 675 point sources, 15,943 AMR grids, and $6.01
\times 10^7 \approx 392^3$ computational cells in this calculation at
this redshift.  On average, $4.6 \times 10^8$ ray segments are traced
each timestep.  The ionised volume fraction is 0.10.  The calculations
are performed on the Nehalam nodes on Pleiades on 2$^n$ cores, where
$n = [2,...,9]$.

Fig. \ref{fig:strong} shows the strong scaling results of this
calculation.  The hydrodynamics and non-equilibrium chemistry routines
scale very well in this range because they depend on local phenomena.
Note that the dominant process is the radiation transport instead of
the hydrodynamics when compared to the weak scaling tests.  The error
bars in the figure represent the deviations across all cores, and it
shows that load balancing becomes an issue at 128 and 256 cores.  This
is an even larger problem for the ray tracing because the grids that
host multiple radiation sources will have significantly more work than
a distant grid.  For example if a few central grids host several point
sources, the exterior grids must wait until those grids are finished
ray tracing in order to receive the rays.  This idle time constitutes
the majority of the time spent in the ray communication routines
despite our efforts (see \S\ref{sec:parallel}) to minimize idle time.
For this reason, the communication of rays does not scale in this
problem.  One solution is to split the AMR grids into smaller blocks
based on the ray tracing work.  This could impact the performance of
the rest of \enzo by increasing the number of boundaries and thus
communication.  Nevertheless in simulations that are dominated by
radiation transport, it will be advantageous to construct such a
scheme to increase the feasibility of running larger problems.

>>>>

In this paper, we present our implementation, \moray, of adaptive ray
tracing \citep{Abel02_RT} in the cosmological hydrodynamics adaptive
mesh refinement (AMR) code, \enzo~\citep{BryanNorman1997, OShea2004}.
The radiation field is coupled to the hydrodynamics solver at small
time-scales, enabling it to study radiation hydrodynamical problems.
We have used this code to investigate the growth of an \hii region
from a 100\Ms~Population III (Pop III) star \citep{Abel07}, the early
stages of reionisation from Pop III stars \citep{Wise08_Reion},
radiative feedback on the formation of high redshift dwarf galaxies
\citep{Wise08_Gal, Wise10_Gal}, ultraviolet radiation escape fractions
from dwarf galaxies before reionisation \citep{Wise09}, negative
radiative feedback from accreting Pop III seed black holes
\citep{Alvarez09}, and radiative feedback in accreting supermassive
black holes \citep[][in prep.]{Kim11}.  We have included \moray~in the
latest public release of \enzo\footnote{http://enzo.googlecode.com},
and it is also coupled with the newly added MHD solver in
\enzo~\citep{Wang09}.

We have structured this paper as follows.  In Section 2, we describe
the mathematical connections between adaptive ray tracing and the
radiative transfer equation.  Furthermore, we detail how physics other
than photo-ionisation and photo-heated are included.  We then derive a
geometric correction factor to any ray tracing method to improve
accuracy.  We end the section by describing a new computational
technique to approximate an optically-thin radiation field with ray
tracing and multiple sources.  In Section 3, we cover the details of
our radiation hydrodynamics implementation in \enzo, specifically (1)
the ray tracing algorithms, (2) coupling with the hydrodynamics
solver, (3) several methods to calculate the radiative transfer
timestep, and (4) our parallelisation strategy.  We present our
results from the RT06 radiative transfer tests in Section 4.
Afterwards in Section 5, we show the results from the RT09 radiation
hydrodynamics tests.  In Section 6, we expand on these tests to
include more dynamical and complex setups to demonstrate the
flexibility and high fidelity of \moray.  Section 7 gives the results
from spatial, angular, frequency, and temporal resolution tests.  In
Section 8, we illustrate the improvements from the geometric
correction factor and our optically-thin approximation.  We also show
the effects of X-ray radiation and radiation pressure in this section.
Finally in Section 9, we demonstrate the parallel scalability of
\moray.  Last Section 10 summarises our method and results.

@@

A ray propagates and splits until 
%
\begin{enumerate}
\item the photon has travelled $c \times dt_P$, where $dt_P$ is the
  radiative transfer timestep,
\item its photon flux is almost fully absorbed ($>99.9\%$) in a single
  cell, which significantly reduces the computational time if the
  radiation volume filling fraction is small,
\item the photon leaves the computational domain with isolated
  boundary conditions, or
\item the photon travels $\sqrt{3}$ of the simulation box length with
  periodic boundary conditions.
\end{enumerate}
%
In the first case, the photon is halted at that position and saved,
where it will be considered in the solution of $I_\nu$ at the next
timestep.  In the next timestep, the photon will encounter a different
hydrodynamical and ionisation state, hence $\kappa$, in its path.
Furthermore any time variations of the luminosities will be retained
in the radiation field.  This is how this method retains the time
derivative of the radiative transfer equation.  The last restriction
prevents our method from considering sources external to the
computational domain, but a uniform radiation background can be used
in conjunction with ray tracing in \moray~that adds the local
radiation field to the background intensity.

@@

Ionisation front velocities can approach significant fractions of the
speed of light in steep density gradients and in the early expansion
of the \hii region.  If the ionisation front position is critical to
the calculation, the radiation transport timestep can be derived from
a non-relativistic estimate of the ionisation front velocity
%
\begin{equation}
  \label{eqn:dt_incident}
  v_{\rm IF}(\mathbf{r}) \approx \frac{F(\mathbf{r})}
  {n_{\rm abs}(\mathbf{r})},
\end{equation}
%
based on the incident radiation field at a particular
position\footnote{See \citet{Shapiro06} for the exact calculation of a
  relativistic ionisation front.  Neglecting relativistic terms do not
  affect the solution because the front velocity is only considered in
  the timestep calculation.}.  Alternatively, the propagation of the
ionisation front can be restricted by limiting the change in specific
intensity $I$ to a safety factor $C_{\rm RT,cfl}$, resulting in a
timestep of
%
\begin{equation}
  \label{eqn:vifront1}
  dt_P = C_{\rm RT,cfl} \frac{I}{|dI/dt|}.
\end{equation}
%
We consider the specific intensity after the ray travels through the
cell, so $I = I_0 \exp(-\tau)$, where $\tau = n_{\rm H} \sigma dl$ is
the optical depth through the cell.  The time derivative of $I$ is
%
\begin{equation}
  \label{eqn:vifront2}
  \frac{dI}{dt} = I_0 \exp(-\tau) (-\dot{n}_{\rm H} \sigma dl),
\end{equation}
%
which can be expressed in terms of local optical depth and neutral
fraction,
\begin{equation}
  \label{eqn:vifront3}
    \frac{dI}{dt} = -I \frac{\tau \dot{n}_{\rm H}}{n_{\rm H}}.
\end{equation}
%
$\dot{n}_{\rm H}$ is computed with the same formula as equation
(\ref{eqn:dtelec}). Substituting in equation (\ref{eqn:vifront1})
gives
%
\begin{equation}
  \label{eqn:vifront4}
  dt_P = C_{\rm RT,cfl} \frac{n_{\rm H}}{\tau \dot{n}_{\rm H}}.
\end{equation}
%
In practice, we have found that a ceiling of 3 can be placed on the
optical depth, so optically thick cells do not create a very small
timestep.  We still find excellent agreement with analytical solutions
with this approximation.  We show the accuracy using this timestep
method in \S\ref{sec:dt_dep}.

@@

In Fig. \ref{fig:test1_ifront}, we show (a) the evolution of the
neutral and ionisation fraction as a function of radius at $t = $ 10,
30, 100, and 500 Myr, and (b) the growth of the ionisation front
radius as a function of time and its ratio with the analytical
Str\"{o}mgren radius [equation (\ref{eqn:rstr})].  The ionisation
front has a width of $\sim0.7$ kpc, which is in agreement with the
inherent thickness of $\sim18 \lambda_{\rm mfp} = 0.74$ kpc, given a
13.6 eV mono-chromatic spectrum.  There are small kinks in the neutral
fraction at 1.5 and 3 kpc that corresponds to artifacts created by the
photon package splitting at these radii.  However these do not affect
the overall solution.  One difference between our results and the
codes presented in RT06 is the increasing neutral fraction outside of
the \hii.  This occurs because the initial ionised fraction and
temperature is set to $1.2 \times 10^{-3}$ and $10^4$ K, which are not
the equilibrium values.  Over the 500 Myr in the calculation, the
neutral fraction increases to 0.2, which is close to its equilibrium
value.  In the right panel of Fig. \ref{fig:test1_ifront}, the
ionisation front radius exceeds $R_s$ by a few percent for most of the
calculation.  This difference happens because the analytical solution
[equation (\ref{eqn:rstr})] assumes the \hii region has a constant
ionised fraction.  The evolution of the ionised fraction as a function
of radius can be analytically calculated \citep[e.g.][]{Osterbrock89,
  Petkova09}, causing the ionisation front radius to be slightly
larger, increasing from 0 to 3\% in the interval 80--350 Myr.  Our
results are in excellent agreement with this more accurate analytical
solution.  In Fig. \ref{fig:test1_HI}, we show a slice of the neutral
fraction through the origin.  Other than the ray splitting artifacts
that generate the plateaus at 1.5 and 3 kpc, one sees spherical
symmetry without any noise in our solution.

@@

\section{Radiation Hydrodynamics Applications}
\label{sec:apps}

We have completed presenting results from the RT06 and RT09 test
suites.  Next we test \moray~with six applications that further
demonstrate its capabilities and accuracy.

\subsection{Application 1. Champagne flow from a dense clump}

\begin{figure*}
  \epsscale{2}
  \plotone{f25}
  \epsscale{1}
  \caption{\label{fig:test8_1} Application 1. (Champagne flow from a dense
    clump).  Slices of density through the initial clump centre in the
    x-y plane (top) and x-z plane (bottom) at $t = 10, 40, 100, 150$
    kyr.  Notice the instabilities that grow from perturbations
    created while the \hii region is contained in the dense
    clump.}
\end{figure*}

@@

\subsection{Application 2. Irradiated Rayleigh-Taylor instability}

Here we combine the classic case of a Rayleigh-Taylor instability and
an expanding \hii region.  The Rayleigh-Taylor instability
occurs when a dense fluid is being supported by a lighter fluid,
initially in hydrostatic equilibrium, in the presence of a constant
acceleration field.  This classic test alone evaluates how subsonic
perturbations evolve.  We consider the case of a single-mode
perturbation.  The system evolves without any radiation until the
perturbation grows considerably and then turn on the radiation source.
These tests demonstrate that \moray~can follow a highly dynamic
system and resolve fine density structures.

@@

\begin{figure}
  %\epsscale{0.75}
  \plotone{f26a}
  \plotone{f26b}
  \plotone{f26c}
  \caption{\label{fig:test9_thick} Application 2. (Irradiated Rayleigh-Taylor
    instability; optically thick case).  Slices at $y=0$ of density
    (top), temperature (middle), and electron fraction (bottom).  The
    source turns on at $t=0$.}
  %\epsscale{1}
\end{figure}

\subsection{Application 3. Photo-evaporation of a blastwave}

\begin{figure*}
  \epsscale{2}
  \plotone{f28}
  \epsscale{1}
  \caption{\label{fig:test10} Application 3 (Photo-evaporation of a
    blastwave).  Slices of density (top) and temperature (bottom) at
    $t = 2.5, 3, 5, 7.5$ Myr in the $x-z$ plane.  As the R-type
    ionisation front propagates through the blastwave centre,
    instabilities grow from the slightly inhomogeneous hot and rarefied
    medium.  Note that the dense shell of the blastwave also creates
    dense inward fingers in the ionisation front shock.}
  %\epsscale{1}
\end{figure*}

@@

\subsection{Application 4. Collimated radiation from a dense clump}

\begin{figure*}
  %\epsscale{1}
  \epsscale{2}
  \plotone{f29}
  \epsscale{1}
  \caption{\label{fig:test11} Application 4 (Collimated radiation from a
    dense clump).  Slices of density (top) and temperature (bottom) at
    $t = 0.1, 3.25, 10.75, 23.25$ Myr.  The conical \hii region
    drives shocks transversely into the overdense sphere and creates
    polar champagne flows.  The ambient medium is heated to $T \sim 3
    \times 10^4$ K as the ionisation front passes the
    constant-pressure cloud surface.  The ionisation front changes
    from D-type to R-type after it enters the ambient medium.}
  %\epsscale{1}
\end{figure*}

@@

We use a domain that is 2 kpc wide and has an ambient medium with
$\rho_0 = 10^{-3}$ \cubecm, $T = 10^4$ K, $x_e = 0.99$.  We place a
dense clump with $\rho/\rho_0 = 100$, $T = 100$ K, $x_e = 10^{-3}$,
and $r = 250$ pc, at the centre of the box.  Radiation is emitted in
two polar cones with $\theta_c = \pi/6$ with 768 (HEALPix level 3)
initial rays, a total luminosity of $10^{49}$ erg s$^{-1}$, and a 17.6
eV mono-chromatic spectrum.  This results in $t_{\rm rec} = 1.22$ Myr
and $R_s = 315$ pc, just outside of the sphere.  The base grid has a
resolution of 64$^3$, and it is refined with the same overdensity
criterion as Application 1.  We run this test for 25 Myr.

\begin{figure}
  %\epsscale{1.15}
  \plottwo{f30a}{f30b}
  \caption{\label{fig:test12_kph} Application 4 (Time variations of the
    source luminosity).  Top: Slice of the photo-ionisation rate
    \kph~through the origin.  The source has a duty cycle of 0.5 Myr,
    and the box has a light crossing time of 3.3 Myr.  The shells of
    high \kph~originate from radiation that was emitted when the
    source was at its peak luminosity, illustrating the
    time-dependence of the radiative transfer equation.  Bottom:
    Radial profile \kph~with the inverse square law overplotted.}
  %\epsscale{1}
\end{figure}

\begin{figure}
  %\epsscale{0.8}
  \plotone{f31}
  \caption{\label{fig:test12_profiles} Application 4 (Time variations of the
    source luminosity).  Radial profiles of (clockwise from upper
    left) density, temperature, pressure, and ionised fraction at $t =
    50, 100, 150, 200, 250$ Myr.  In this problem, the time variations
    in the source have little effect on the overall \hii region
    expansion.  Inside the ionisation front at $t = 50$ Myr, there are
    small density perturbations that are created by the variable
    source that are later smoothed out over a sound crossing time.}
  %\epsscale{1}
\end{figure}

@@

\subsection{Application 5. Time variations of the source luminosity}

\begin{figure*}
  \epsscale{2}
  \plotone{f32}
  \epsscale{1}
  \caption{\label{fig:test14_1} Application 5. (\hii region with MHD).
    Left to right: slices of density at $t = 0.18, 0.53, 1.58$ Myr in
    the x-y plane.  The streamlines show the magnetic field.}
\end{figure*}

\begin{figure*}
  \epsscale{2}
  \plotone{f33}
  \epsscale{1}
  \caption{\label{fig:test14_2} Application 5. (\hii region with
    MHD). Slices of density (top) and the x-component of the magnetic
    field (bottom) in the y-z plane at $t = 0.18, 0.53, 1.58$ Myr
    (left to right).}
\end{figure*}

@@

\subsection{Application 6. \hii region with MHD}

@@

Last we demonstrate the parallel performance of \moray~in weak and
strong scaling tests.  For large simulations to consider radiative
transfer, it is imperative that the code scales to large number of
processors.

@@

Weak scaling tests demonstrates how the code scales with the number of
processors with a constant amount of work per processor.  Here we
construct a test problem with a $64^3$ block per core.  The grid is
not adaptively refined.  The physical setup of the problem is nearly
the same as Test 5 with a uniform density $\rho = 10^{-3} \;\cubecm$
and initial temperature $T = 100$ K.  Each block has the same size of
15 kpc as Test 5.  At the centre of each grid, there exists a
radiation source with a luminosity $L = 5 \times 10^{48}$ ph s$^{-1}$
and a 17 eV mono-chromatic spectrum.  The problem is run for 250 Myr.
We run this test with $N_p = 2^n$ cores with $n = [0,1 \dots 10,11]$.
The domain has $(N_x, N_y, N_z)$ blocks that is determined with the
MPI routine \texttt{MPI\_Dims\_create}.  For example with $n = 7$, the
problem is decomposed into $(N_x, N_y, N_z) = (4,4,8)$ blocks,
producing a $256 \times 256 \times 512$ grid.  We have run these on
the original (Harpertown CPUs) nodes of the NASA NAS machine,
Pleiades, with 8 cores per node.  Fig. \ref{fig:weak} shows the
performance timings of various parts of the code.  From one to two
cores, the total time only increases by $\sim$1\% due to the overhead
associated with the inter-processor message passing.  From two to
eight cores, the performance decreases in the hydrodynamics solver,
chemistry and energy solver, and obtaining the hydrodynamic boundary
conditions because of cache locality problems of the data being passed
to the CPU.  This occurs because of the CPU architecture, specifically
the L1 cache and core connectivity.  We see less of a penalty in newer
processors, e.g. Intel Nehalam and Westmere CPUs.  Above eight
processes, these routines exhibit near perfect weak scaling to 2048
cores.  Unfortunately there exists a $N_p^{1.5}$ dependence in the ray
communication routines.  It becomes the dominant process above 512
processes.  We are actively pursing a solution to this scaling
problem.  The other parts of the code exhibit excellent weak scaling.
Overall it scales already well to 512 processes, and there is room to
enhance the weak scalability of \moray~in the near future.

@@

Strong scaling tests shows how the problem scales with the number of
processors for the same problem.  The overhead associated with the
structured AMR framework in \enzo~can limit the strong scalability.
One key property of strong scaling is that each processor must have
sufficient work to compute, compared to the communication involved.
In our experience, non-AMR calculations exhibit much better strong
scaling than AMR ones because of reduced inter-processor
communication.  We use an AMR simulation to demonstrate the scalability
of \moray~in a demanding, research application.  Here we use a
small-box reionisation calculation with $L_{\rm box} = 3$ Mpc/h, a
resolution of $256^3$, and six levels of refinement.  We measure the
time spent on the hydrodynamics, non-equilibrium chemistry, ray
tracing, and radiation transport communication in a timestep lasting 1
Myr, at $z=10$.  There are nine radiative transfer timesteps in this
period.  The box has 675 point sources, 15,943 AMR grids, and $6.01
\times 10^7 \approx 392^3$ computational cells in this calculation at
this redshift.  On average, $4.6 \times 10^8$ ray segments are traced
each timestep.  The ionised volume fraction is 0.10.  The calculations
are performed on the Nehalam nodes on Pleiades on 2$^n$ cores, where
$n = [2,...,9]$.

Fig. \ref{fig:strong} shows the strong scaling results of this
calculation.  The hydrodynamics and non-equilibrium chemistry routines
scale very well in this range because they depend on local phenomena.
Note that the dominant process is the radiation transport instead of
the hydrodynamics when compared to the weak scaling tests.  The error
bars in the figure represent the deviations across all cores, and it
shows that load balancing becomes an issue at 128 and 256 cores.  This
is an even larger problem for the ray tracing because the grids that
host multiple radiation sources will have significantly more work than
a distant grid.  For example if a few central grids host several point
sources, the exterior grids must wait until those grids are finished
ray tracing in order to receive the rays.  This idle time constitutes
the majority of the time spent in the ray communication routines
despite our efforts (see \S\ref{sec:parallel}) to minimize idle time.
For this reason, the communication of rays does not scale in this
problem.  One solution is to split the AMR grids into smaller blocks
based on the ray tracing work.  This could impact the performance of
the rest of \enzo~by increasing the number of boundaries and thus
communication.  Nevertheless in simulations that are dominated by
radiation transport, it will be advantageous to construct such a
scheme to increase the feasibility of running larger problems.
